Predicting Severity of _C. difficile_ Infections from the Taxonomic Composition of the Gut Microbiome

### 2023-06-07 paper/paper.qmd
Minor consistency & typo fixes

Since the data are highly imbalanced with only a small proportion of CDI cases
having a severe outcome, evaluating the trade-off between precision and recall
is more informative than the receiver-operator characteristic because precision
and recall do not consider true negatives, which may overinflate the AUROC.
However, unlike for AUROC, the baseline for the area under the precision-recall
curve depends on the proportion of positive outcomes (i.e. severe cases) in the
data, which vary across these severity definitions.
To allow comparison of precision across datasets with different proportions of
positives, @wu_improved_2021 introduced the concept of balanced precision, a
transformation of precision based on Bayes' theorem that represents the
precision that would have been expected if the proportion of positives were
balanced at 0.5.
Reporting the area under the balanced precision-recall curve (AUBPRC) allows us
to compare the trade-off between precision and recall for our different severity
defintions.
The test set median AUBPRCs from the full datasets followed a similar pattern as 
the test set AUROCs with
`r perf_medians$aubprc_full_idsa` for IDSA severity,
`r perf_medians$aubprc_full_allcause` for all-cause severity,
`r perf_medians$aubprc_full_attrib` for attributable severity, and
`r perf_medians$aubprc_full_pragmatic` for pragmatic severity.
The AUBPRCs were significantly different from each other (P < 0.05) for
each pair of severity definitions except for attributable vs all-cause.
We plotted the balanced precision-recall curve and found that the IDSA definition
outperformed all other models at very low recall values, but the others outperform
IDSA at all other points of the curve (@fig-performance C).
The 95% confidence intervals overlapped the baseline AUROC and AUBPRC for
the attributable severity models, while all others did not overlap the baseline.

While it is advantageous to use as much data as available to train the best
models possible, comparing performances of models trained on different subsets
of the data is not entirely fair.
To enable fair comparisons of the model performances across different severity
definitions, we also selected the intersection of samples (n=`r n_cases_int`)
that had labels for all four severity definitions and repeated the model
training and evaluation process on this intersection dataset.
The attributable definition is exactly the same as the pragmatic definition for
the intersection dataset, as we defined pragmatic severity to use the
attributable label when available.
The performance results on the intersection dataset are shown in the right
facets of each panel of @fig-performance.

@@

As with the full datasets, the AUROCs of the training sets and test sets were
similar within each severity definition.
The median test set AUROCs were 
`r perf_medians$auroc_int_idsa` for IDSA severity,
`r perf_medians$auroc_int_allcause` for all-cause severity,
`r perf_medians$auroc_int_attrib` and for attributable severity.
The AUROCs on the intersection dataset were significantly different for
all-cause vs attributable and all-cause vs IDSA severity (P < 0.05), but not for
IDSA vs attributable severity (P > 0.05).
The median test set AUBPRCs were
`r perf_medians$aubprc_int_idsa` for IDSA severity,
`r perf_medians$aubprc_int_allcause` for all-cause severity,
`r perf_medians$aubprc_int_attrib` and for attributable severity.
Just as with the AUROCs, the AUBPRCs were significantly different for
all-cause vs attributable and all-cause vs IDSA severity (P < 0.05), but not for
IDSA vs attributable severity (P > 0.05).
For all severity definitions, performance dropped between the full dataset and
the intersection dataset since fewer samples are available, but this effect is
least dramatic for IDSA severity as the full and intersection datasets are more
similar for this definition (@tbl-counts B).
The 95% confidence interval overlaps with the baseline for both AUROC and AUBPRC
for all definitions on the intersection dataset except for IDSA severity.

@@

We trained ML models based on gut microbial communities on the day of CDI
diagnosis to predict CDI severity according to four different severity definitions.
The purpose of the full datasets was to train the best models possible given the
constraints, while using the intersection dataset allows for comparing severity
definitions.
We found that models predicting pragmatic severity with as much data as
available performed best, while models classifying IDSA severity outperformed
the all-cause and attributable definitions only with the inersection.
Performance dropped substantially when reducing to the intersection dataset for
all definitions, likely due to the particularly imbalanced nature of the
all-cause and attributable definitions.
These results demonstrate the importance of using as many samples as possible when
data are sparse and the outcome is low prevalence, as well as the need to
incorporate physician's expertise when possible.

Permutation feature importance revealed patterns of important bacteria that
concord with the literature.
Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and
severity has been well-documented in prior studies, thus its importance and
increase in abundance for severe cases is not surprising
[@schubert_microbiome_2014;
@antharam_intestinal_2013;
@berkell_microbiota-based_2021;
@lesniak_gut_2022].
For many of the top OTUs, there is a wide range in importance.
Notably, the OTU represented by _Pseudomonas_ had wide variance in importance
for the full dataset in models predicting attributable severity, with the
maximum point more important than any other OTU yet a minimum below zero.
However, for the intersection dataset, this OTU was removed due to having
near-zero variance.
The presence of _Pseudomonas_ was thus informative in a small number of 
patient samples, but not in others, and these samples were lost in the
intersection dataset.
Overall the abundance data are patchy, as these patients were likely all on
antibiotics for unrelated infections prior to CDI onset.
A limitation of permutation importance is that the contribution of each feature
is considered in isolation, but members of microbial communities interact and
compete with each other, thus these complicated relationships are not well 
captured by permutation importance.

@@

However, it is not enough for models to perform well to justify deploying them
in a clinical setting; benefit over current practices must be shown [@wiens_no_2019].
Although no known treatment options have been shown to reduce the risk of severe
CDI outcomes, fidaxomicin is promising due to its improved time to resolution
and reduced recurrence.
Despite its increased cost, fidaxomicin is also attractive as a preferential
antibiotic option as vancomycin-resistant _Enterococcus_ is on the rise and
enterococci are known to worsen CDI
[@poduval_clostridium_2000;@smith_enterococci_2022].
We extended our analysis of clinical value to incorporate the number needed to
treat for fidaxomicin alongside the predictive models in order to calculate the
number needed to benefit.
The NNB contextualizes model performance within clinical reality, as it combines
both model performance and treatment effectiveness [@liu_number_2019].
A more robust analysis of clinical value would further consider the cost of
treatment options versus the savings of averting severe outcomes, as economic
disparities are a major barrier to treatment in the US [@johnson_clinical_2021].
Cost-benefit analyses based on clinical trial data have reported that
fidaxomicin may be as cost-effective as vancomycin as a treatment for initial
CDI cases, largely due to the reduced risk of recurrence
[@jiang_budget_2022;@reveles_fidaxomicin_2017].
While our analysis of clinical value is only a proof-of-concept, if evidence
emerges that new or existing treatments significantly reduce the risk of severe
CDI, our results can be incorporated into future considerations of whether to
build severity prediction models and what features should be incorporated.
In practice, EHR-based models are less costly to deploy than OTU-based models
and do not require additional clinical sample collection.
Amplicon sequencing is not typically performed for CDI patients, however,
routinely profiling the microbial communities of CDI patients could be justified
if models that incorporate microbial features were shown to improve patient
outcomes.

>>>>

Since the data are highly imbalanced with only a small proportion of CDI cases
having a severe outcome, evaluating the trade-off between precision and recall
is more informative than the receiver-operator characteristic because precision
and recall do not consider true negatives, which may overinflate the AUROC.
However, unlike for AUROC, the baseline for the area under the precision-recall
curve depends on the proportion of positive outcomes (i.e. severe cases) in the
data, which vary across these severity definitions.
To allow comparison of precision across datasets with different proportions of
positives, @wu_improved_2021 introduced the concept of balanced precision, a
transformation of precision based on Bayes' theorem that represents the
precision that would have been expected if the proportion of positives were
balanced at 0.5.
Reporting the area under the balanced precision-recall curve (AUBPRC) allows us
to compare the trade-off between precision and recall for our different severity
defintions.
The test set median AUBPRCs from the full datasets followed a similar pattern as 
the test set AUROCs with
`r perf_medians$aubprc_full_idsa` for IDSA severity,
`r perf_medians$aubprc_full_allcause` for all-cause severity,
`r perf_medians$aubprc_full_attrib` for attributable severity, and
`r perf_medians$aubprc_full_pragmatic` for pragmatic severity.
The AUBPRCs were significantly different from each other (P < 0.05) for
each pair of severity definitions except for attributable versus all-cause.
We plotted the balanced precision-recall curve and found that the IDSA definition
outperformed all other models at very low recall values, but the others outperform
IDSA at all other points of the curve (@fig-performance C).
The 95% confidence intervals overlapped the baseline AUROC and AUBPRC for
the attributable severity models, while all others did not overlap the baseline.

While it is advantageous to use as much data as available to train the best
models possible, comparing performances of models trained on different subsets
of the data is not entirely fair.
To enable fair comparisons of the model performances across different severity
definitions, we also selected the intersection of samples (n=`r n_cases_int`)
that had labels for all four severity definitions and repeated the model
training and evaluation process on this intersection dataset.
The attributable definition is exactly the same as the pragmatic definition for
the intersection dataset, as we defined pragmatic severity to use the
attributable definition when available.
The performance results on the intersection dataset are shown in the right
facets of each panel of @fig-performance.

@@

As with the full datasets, the AUROCs of the training sets and test sets were
similar within each severity definition.
The median test set AUROCs were 
`r perf_medians$auroc_int_idsa` for IDSA severity,
`r perf_medians$auroc_int_allcause` for all-cause severity,
`r perf_medians$auroc_int_attrib` and for attributable severity.
The AUROCs on the intersection dataset were significantly different for
all-cause versus attributable and all-cause versus IDSA severity (P < 0.05), but not for
IDSA versus attributable severity (P > 0.05).
The median test set AUBPRCs were
`r perf_medians$aubprc_int_idsa` for IDSA severity,
`r perf_medians$aubprc_int_allcause` for all-cause severity,
`r perf_medians$aubprc_int_attrib` and for attributable severity.
Just as with the AUROCs, the AUBPRCs were significantly different for
all-cause versus attributable and all-cause versus IDSA severity (P < 0.05), but not for
IDSA versus attributable severity (P > 0.05).
For all severity definitions, performance dropped between the full dataset and
the intersection dataset since fewer samples are available, but this effect is
least dramatic for IDSA severity as the full and intersection datasets are more
similar for this definition (@tbl-counts B).
The 95% confidence interval overlaps with the baseline for both AUROC and AUBPRC
for all definitions on the intersection dataset except for IDSA severity.

@@

We trained ML models based on gut microbial communities on the day of CDI
diagnosis to predict CDI severity according to four different severity definitions.
The purpose of the full datasets was to train the best models possible given the
constraints, while using the intersection dataset allows for comparing severity
definitions.
We found that models predicting pragmatic severity with as much data as
available performed best, while models classifying IDSA severity outperformed
the all-cause and attributable definitions only with the intersection.
Performance dropped substantially when reducing to the intersection dataset for
all definitions, likely due to the particularly imbalanced nature of the
all-cause and attributable definitions.
These results demonstrate the importance of using as many samples as possible when
data are sparse and the outcome is low prevalence, as well as the need to
incorporate physician's expertise when possible.

Permutation feature importance revealed patterns of important bacteria that
concord with the literature.
Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and
severity have been well-documented in prior studies, thus their importance and
increase in abundance for severe cases is not surprising
[@schubert_microbiome_2014;
@antharam_intestinal_2013;
@berkell_microbiota-based_2021;
@lesniak_gut_2022].
For many of the top OTUs, there is a wide range in importance.
Notably, the OTU represented by _Pseudomonas_ had wide variance in importance
for the full dataset in models predicting attributable severity, with the
maximum point more important than any other OTU yet a minimum below zero.
However, for the intersection dataset, this OTU was removed due to having
near-zero variance.
The presence of _Pseudomonas_ was thus informative in a small number of 
patient samples, but not in others, and these samples were lost in the
intersection dataset.
Overall the abundance data are patchy, as these patients were likely all taking
antibiotics for unrelated infections prior to CDI onset.
A limitation of permutation importance is that the contribution of each feature
is considered in isolation, but members of microbial communities interact and
compete with each other, thus these complicated relationships are not well 
captured by permutation importance.

@@

However, it is not enough for models to perform well to justify deploying them
in a clinical setting; benefit over current practices must be shown [@wiens_no_2019].
Although no known treatment options have been shown to reduce the risk of severe
CDI outcomes, fidaxomicin is promising due to its improved time to resolution
and reduced recurrence.
Despite its increased cost, fidaxomicin is also attractive as a preferential
antibiotic option as vancomycin-resistant _Enterococcus_ is on the rise and
enterococci are known to worsen CDI
[@poduval_clostridium_2000;@smith_enterococci_2022].
We extended our analysis of clinical value to incorporate the number needed to
treat for fidaxomicin alongside the predictive models in order to calculate the
number needed to benefit.
The NNB contextualizes model performance within clinical reality, as it combines
both model performance and treatment effectiveness [@liu_number_2019].
A more robust analysis of clinical value would further consider the cost of
treatment options versus the savings of averting severe outcomes across a range
of decision thresholds, as economic disparities are a major barrier to treatment
in the US [@johnson_clinical_2021].
Cost-benefit analyses based on clinical trial data have reported that
fidaxomicin may be as cost-effective as vancomycin as a treatment for initial
CDI cases, largely due to the reduced risk of recurrence
[@jiang_budget_2022;@reveles_fidaxomicin_2017].
While our analysis of clinical value is only a proof-of-concept, if evidence
emerges that new or existing treatments significantly reduce the risk of severe
CDI, our results can be incorporated into future considerations of whether to
build severity prediction models and what features should be incorporated.
In practice, EHR-based models are less costly to deploy than OTU-based models
and do not require additional clinical sample collection.
Amplicon sequencing is not typically performed for CDI patients, however,
routinely profiling the microbial communities of CDI patients could be justified
if models that incorporate microbial features were shown to improve patient
outcomes.
